---
title: "Seal Tag Data Visualisation Tutorial"
author: "Jake Weis, Denisse Fierro Arcos"
format: gfm
---

## Introduction
This tutorial was originally developed by Jake Weis and updated by Denisse Fierro Arcos.  

The main aim of this tutorial is to show students how to visualise oceanographic data collected by seal-borne tags. These tags measure various ocean properties, such as temperature, salinity and fluorescence, at different depths and locations. Data from these tags are often made available in netCDF format, which is a commonly used format for oceanographic data.  

By the end of this tutorial, you will be able to:  
1. Load and examine netCDF files  
2. Create maps showing spatial distribution of the seal tag data  
3. Create time series plots of sea surface parameters  
4. Create "pressure section" plots showing how ocean properties change across depths, time and space
5. Interpret oceanographic data with respect to seal foraging behaviour

## Loading relevant `R` packages
To access all functions needed in this tutorial, you will need to load any relevant packages first. You can load packages using the `library` function as shown below. Make sure you read the tutorial instructions and load all relevant packages here.

```{r, warning = FALSE, output = FALSE}
library(terra)

# Load all other relevant packages here
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(janitor)
library(ncdf4)
library(metR)
library(ggplot2)
library(cmocean)
library(patchwork)
```

## PART A: LOADING SEAL TAG DATA FROM NETCDF FILES
First, you will learn how to open and explore the netCDF files containing the seal tag data. This will help you understand how the data is structured and how to extract the variables you need for analysis.  

Since this tutorial is stored as an R project, you can provide relative file paths. That is, the path should be provided in relation to your current directory. If you don't know what your current directory is, you can use the `getwd()` function.  
  
```{r, eval = FALSE}
getwd()
```

All netCDF files used in this tutorial are stored in the `data` folder, which means you can use the `list.files` functions to get the filepaths for all netCDF files. This way you can avoid typing out the filepath for each file of interest.  

```{r}
# Get a list of netCDF files stored in the data folder
ras_filepaths <- list.files("../data", pattern = ".nc$", full.names = TRUE)

# Check the files available
ras_filepaths
```

### Using `terra` to load netCDF files as gridded (cube) data
The `terra` package for `R` is designed to performed spatial data analysis using vector (e.g., lines, points, polygons) and raster (gridded) data. It is one of the most commonly used packages for spatial analysis.

You can choose whichever file to explore its contents, but you will need to create plots for all files.  

```{r}
# Choose a netCDF file from the data folder and load it here
ras <- rast(ras_filepaths[1])

# Check its contents by calling the variable you assigned the netCDF file to
ras
```

We can see that this file contains multiple variables (e.g., depth, chlorophyll fluorescence, etc.). We can also see the units for each one of these variables, the spatial extent of the data, its resolution, but it does not have information about a coordinate reference system (`coord. ref`) or time.  

The `terra` package has multiple functions that allow us to further query the metadata (i.e., information about the file) contained in the netCDF file. Below, you will find a few functions listed. Check out the help to find out how to use them, but the function name is self-explanatory.  

```{r}
varnames(ras)
```

`terra` also gives us the option to look at the long names for the variables contained in the file. These will include a better description of what information each variable contains.  

```{r}
longnames(ras)
```

You may notice that the long names for the fluorescence variables are all the same, so we cannot tell exactly what the difference is between them. For now, let's move onto checking the temporal range covered by the data.  

```{r}
# Checking time
time(ras)
```

`terra` cannot identify the time in this file, which is an issue when attempting to create time series plots. Finally, let's check the coordinate reference system (CRS).  

```{r}
# Checking coordinate reference system (CRS)
crs(ras)
```

Again, `terra` does not return anything for the CRS, which is problematic if we want to create maps. We can attempt to access all metadata from the file to see if this information was included.

```{r}
# We can also check all metadata
meta(ras)
```

It appears `terra` is not reading the complete file metadata, which is not ideal. You may have also noticed at the beginning of this section that the spatial extent of this file goes well beyond the possible values for latitude and longitude (in decimal degrees). We can check the spatial extent using the `ext` function from the `terra` package, which will return the minimum and maximum values along the `x` and `y` dimensions of the raster file. Try it below.  

```{r}
ext(ras)
```

As you can see, the maximum value for the `x` dimension goes well beyond $\pm 180 ^\circ$ or $360 ^\circ$, and the maximum latitude is also beyond $\pm 90 ^\circ$. So, what is happening? Can we trust this information? Probably not.  

Can we do anything in this case? Yes, luckily, there are plenty of other packages available in `R` that allow us to load and analyse gridded data. Below, we will use two of these packages: `ncdf4` and `metR`.  

### Using `ncdf4` and `metR` to load netCDF files as gridded data
We will load and explore the content of netCDF files using the `ncdf4` package.  

```{r}
nc_file <- nc_open(ras_filepaths[1])
nc_file
```

As you can see, we get much detailed metadata using this package. We can even see a full description for the variables with names starting with `FLUO`, which means we can now understand how they all differ. We have all the extra information we need. We will store some of this metadata though because it may be useful when processing data. For now, we will store metadata for temperature, salinity, fluorescence, and time variables, as well as global attributes. For the fluorescence we will choose the the dark-, NPQ-corrected fluorescence, and you can find more details about the reasons why in Section B1.  

```{r}
time_attrs <- ncatt_get(nc_file, "JULD")
temp_attrs <- ncatt_get(nc_file, "TEMP")
salt_attrs <- ncatt_get(nc_file, "PSAL")
fluo_attrs <- ncatt_get(nc_file, "FLUO_DRK_NPQ")
global_attrs <- ncatt_get(nc_file, 0)
```

You can check the metadata by calling the variable of interest.  

```{r}
fluo_attrs
```

Now that we have every we need, we can close the connection to the netCDF file. But if you later realise you forgot something, you can either load the file and extract metadata as we have done above or simply refer to the printed metadata above.  

```{r}
nc_close(nc_file)
```

We will now use the `metR` package to load the netCDF file as a data frame as this facilitates data analysis in `R`. 

```{r}
# Load the file
df <- ReadNetCDF(ras_filepaths[1])

# You can check the first few rows of this file
head(df)
```

You can check that all variables included in the metadata are included in the data frame above. However, we will clean up this data frame a little. First, we will use the `janitor` package to change the column names to lowercase so it is easy to refer to them. Afterwards, we will need to change the `JULD` column to a date as it currently is in number of days. How do I know this? I checked the time metadata we stored above. Print it below and see for yourself.  

```{r}
time_attrs
```

As you can see in the units, `JULD` is in "`{r} print(time_attrs$units)`". We will store the origin date so we can create a `time` column in our data frame.

```{r}
# Get start date from metadata. We will remove all letters
time_origin <- str_remove_all(time_attrs$units, "[A-z]") |> 
  # Then we will remove whitespaces at the beginnind and end
  str_trim()

time_origin
```

Let's clean our data frame.  

```{r}
df <- df |> 
  clean_names() |> 
  # Remember to check the origin date above to get the correct time
  mutate(time = as_date(juld, origin = time_origin), .after = "juld")

# Check results
head(df)
```

Our data frame is looking much better and it is now ready for data analysis or visual exploration.  

## PART B: CREATE A MAP OF SEAL LOCATIONS
In this section, you will create a map of the seal tag profile locations. You will colour the points based on ocean properties. This type of map will show you not only where the seal traveled during its foraging trip, but also give you an idea of the changes in water properties it experienced along the way.  

### B1: Calculate mean sea surface temperature (SST), salinity (SSS) and fluorescence (SSF)
To calculate the mean values for these three variables, you will use data for the top 20 m, which you can find in the data frame we loaded in the step prior.  
  
You probably already noticed that there are multiple fluorescence variables. For this exercise, we will be using the dark-, NPQ-corrected fluorescence. NPQ stands for "non-photochemical-quenching", a mechanism that acts as a sort of sunblock for phytoplankton, reducing their fluorescence during daytime. NPQ allows cells to dissipate excess energy as heat to prevent damage to the cell's reaction centres, but this inhibits photosynthesis and decreases fluorescence. NPQ-corrected fluorescence will therefore exclude any low fluorescence values linked to high sunlight level and that are not reflective of real variations in chlorophyll-a concentration and phytoplankton biomass.  
  
```{r}
df_mean20 <- df |> 
  # Start by selecting any relevant columns
  select(latitude, longitude, depth, time, temp, psal, fluo_drk_npq) |> 
  # Now filter values for the top 20 m
  filter(depth >= -20) |>
  # Finally calculate the mean for the variables of interest
  summarise(across(temp:fluo_drk_npq, ~ mean(.x, na.rm = TRUE)), 
            .by = c(latitude, longitude, time))

head(df_mean20)
```

### B2: Map seal tag profiles coloured by SST
First, we will get a world map layer to use as a base map in our maps.  

```{r}
world <- map_data("world")
```

Now we will create a map with polar stereographic projection using the seal tag data, which is distributed around the Southern Ocean. Remember that we will colour the tracks by the sea surface temperature each animal experienced during their journey.  

```{r}
profile_map <- ggplot() +
  # Adding world base layer using geom_polygon(). Note that "fill" and "color" 
  # control the color of the continents and borders, respectively.
  geom_polygon(data = world,
               aes(x = long, y = lat, group = group),
               fill = "lightgray", color = "black", linewidth = 0.2) +
  # Adding points for profile locations (lat/lon coordinates) and colouring them
  # by sea surface temperature.
  # The coordinates are stored in the data frame we processed above. Make sure
  # you use the correct names and axes for the lat/lon variables, and provide a
  # variable in the "colour" option for dynamic colouring.
  geom_point(data = df_mean20, aes(x = longitude, y = latitude, colour = temp),
             size = 1.5, alpha = 0.7) +
  # Use an appropriate color scale. CMOCEAN is a great resource for 
  # colorblind-friendly color palettes. You can find the palettes to choose 
  # from and their names here:
  # https://cran.r-project.org/web/packages/cmocean/vignettes/cmocean.html
  scale_color_cmocean("SST (deg C)", name = "thermal") +
  # Setting polar projection ("ortho") centered on South Pole (-90˚).
  # Note: To rotate the map, change the second "orientation" value.
  coord_map("ortho", orientation = c(-90, 0, 0)) +
  # Adding labels
  labs(title = paste0("Seal Tag Profile Locations (Platform code: ",
                      global_attrs$platform_code, ")"),
       subtitle = paste0("Total profiles: ", max(df$n_prof))) +
  # Changing x and y axis breaks
  scale_x_continuous(breaks = seq(-180, 180, 30)) +
  scale_y_continuous(breaks = seq(-90, 90, 15)) +
  # Making the map a little nicer
  theme_minimal() +
  theme(panel.grid = element_line(color = "lightgray", linewidth = 0.2),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        legend.position = "right", legend.title = element_text(hjust = 0.5),
        axis.title = element_blank(), axis.text = element_blank(), 
        plot.title = element_text(hjust = 0.5))

# Checking the result
profile_map
```

We can also save the map as an image to our computer.  

```{r}
# We can create a figure directory
figure_folder <- "../figures"

if(!dir.exists(figure_folder)){
  dir.create(figure_folder)
}

# Now we can save the figure into the correct directory
ggsave(file.path(figure_folder, "profile_map_temp.png"), profile_map,
       width = 12, height = 8, dpi = 300)
```

Now you can try creating maps using other variables to colour the tracks. You can do this by adjusting the code shown in this section (B2).  

Make sure you also save the map to your computer, but don't forget to change the filename so the figure is not overwritten. Finally, feel free to explore other colour palettes that may fit the variable being shown a little better (a green palette may be better for fluorescence, for example).  

### B3: Map the seal tag profiles coloured by SSS

```{r}
# Write your code here
ggplot() +
  # Adding world base layer using geom_polygon(). Note that "fill" and "color" 
  # control the color of the continents and borders, respectively.
  geom_polygon(data = world,
               aes(x = long, y = lat, group = group),
               fill = "lightgray", color = "black", linewidth = 0.2) +
  # Adding points for profile locations (lat/lon coordinates) and colouring them
  # by sea surface temperature.
  # The coordinates are stored in the data frame we processed above. Make sure
  # you use the correct names and axes for the lat/lon variables, and provide a
  # variable in the "colour" option for dynamic colouring.
  geom_point(data = df_mean20, aes(x = longitude, y = latitude, colour = psal),
             size = 1.5, alpha = 0.7) +
  # Use an appropriate color scale. CMOCEAN is a great resource for 
  # colorblind-friendly color palettes. You can find the palettes to choose 
  # from and their names here:
  # https://cran.r-project.org/web/packages/cmocean/vignettes/cmocean.html
  scale_color_cmocean("SSS (psu)", name = "haline") +
  # Setting polar projection ("ortho") centered on South Pole (-90˚).
  # Note: To rotate the map, change the second "orientation" value.
  coord_map("ortho", orientation = c(-90, 0, 0)) +
  # Adding labels
  labs(title = paste0("Seal Tag Profile Locations (Platform code: ",
                      global_attrs$platform_code, ")"),
       subtitle = paste0("Total profiles: ", max(df$n_prof))) +
  # Changing x and y axis breaks
  scale_x_continuous(breaks = seq(-180, 180, 30)) +
  scale_y_continuous(breaks = seq(-90, 90, 15)) +
  # Making the map a little nicer
  theme_minimal() +
  theme(panel.grid = element_line(color = "lightgray", linewidth = 0.2),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        legend.position = "right", legend.title = element_text(hjust = 0.5),
        axis.title = element_blank(), axis.text = element_blank(), 
        plot.title = element_text(hjust = 0.5))
```


### B4: Map the seal tag profiles coloured by SSF

```{r}
# Write your code here
ggplot() +
  # Adding world base layer using geom_polygon(). Note that "fill" and "color" 
  # control the color of the continents and borders, respectively.
  geom_polygon(data = world,
               aes(x = long, y = lat, group = group),
               fill = "lightgray", color = "black", linewidth = 0.2) +
  # Adding points for profile locations (lat/lon coordinates) and colouring them
  # by sea surface temperature.
  # The coordinates are stored in the data frame we processed above. Make sure
  # you use the correct names and axes for the lat/lon variables, and provide a
  # variable in the "colour" option for dynamic colouring.
  geom_point(data = df_mean20, aes(x = longitude, y = latitude, 
                                   colour = fluo_drk_npq),
             size = 1.5, alpha = 0.7) +
  # Use an appropriate color scale. CMOCEAN is a great resource for 
  # colorblind-friendly color palettes. You can find the palettes to choose 
  # from and their names here:
  # https://cran.r-project.org/web/packages/cmocean/vignettes/cmocean.html
  scale_color_cmocean(bquote(SSF~(mg%.%m^-3)), name = "algae") +
  # Setting polar projection ("ortho") centered on South Pole (-90˚).
  # Note: To rotate the map, change the second "orientation" value.
  coord_map("ortho", orientation = c(-90, 0, 0)) +
  # Adding labels
  labs(title = paste0("Seal Tag Profile Locations (Platform code: ",
                      global_attrs$platform_code, ")"),
       subtitle = paste0("Total profiles: ", max(df$n_prof))) +
  # Changing x and y axis breaks
  scale_x_continuous(breaks = seq(-180, 180, 30)) +
  scale_y_continuous(breaks = seq(-90, 90, 15)) +
  # Making the map a little nicer
  theme_minimal() +
  theme(panel.grid = element_line(color = "lightgray", linewidth = 0.2),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        legend.position = "right", legend.title = element_text(hjust = 0.5),
        axis.title = element_blank(), axis.text = element_blank(), 
        plot.title = element_text(hjust = 0.5))
```

## PART C: TIME SERIES OF SEA SURFACE PROPERTIES
Time series plots usually provide a clearer view of how a variable changes over time (and space). You will now create time series plots for sea surface temperature, salinity, and fluorescence. These plots will show how the seal's environment changed during its foraging trip.

### C1: SST time series
Create a line plot showing how SST changed over time. 

```{r}
sst_timeseries <- df_mean20 |>
  ggplot(aes(x = time, y = temp)) +
  # Creating a line plot
  geom_line(color = "red", linewidth = 1) +
  # Adding points to show individual measurements
  geom_point(color = "darkred", size = 1.5, alpha = 0.7) +
  # Creating date tick labels formatted as "YYYY-MM-DD"
  scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 month") +
  # Adding labels to plot
  labs(title = paste0("Mean ocean temperature over top 20 m (deg C) during ", 
                      "seal tagging period"),
       y = "SST (deg C)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"))

# Checking results
sst_timeseries
```

Finally, we can save the timeseries as an image to our computer.  

```{r}
ggsave(file.path(figure_folder, "sst_timeseries.png"), sst_timeseries,
       width = 10, height = 6, dpi = 300)
```

Now it is your turn to create timeseries plots for salinity and fluorescence. Don't forget you can use the code above as an example.  

### C2: SSS time series 
```{r}
# Write your code here
sss_timeseries <- df_mean20 |>
  ggplot(aes(x = time, y = psal)) +
  # Creating a line plot
  geom_line(color = "red", linewidth = 1) +
  # Adding points to show individual measurements
  geom_point(color = "darkred", size = 1.5, alpha = 0.7) +
  # Creating date tick labels formatted as "YYYY-MM-DD"
  scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 month") +
  # Adding labels to plot
  labs(title = paste0("Mean ocean salinity over top 20 m (deg C) during ", 
                      "seal tagging period"),
       y = paste0("SSS (", salt_attrs$units, ")")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"))

# Checking results
sss_timeseries
```

### C3: SSF time series
```{r}
# Write your code here
ssf_timeseries <- df_mean20 |>
  ggplot(aes(x = time, y = fluo_drk_npq)) +
  # Creating a line plot
  geom_line(color = "red", linewidth = 1) +
  # Adding points to show individual measurements
  geom_point(color = "darkred", size = 1.5, alpha = 0.7) +
  # Creating date tick labels formatted as "YYYY-MM-DD"
  scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 month") +
  # Adding labels to plot
  labs(title = paste0("Mean ocean fluorescence over top 20 m (deg C) during ", 
                      "seal tagging period"),
       y = bquote(SSF~(mg%.%m^-3))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"))

# Checking results
ssf_timeseries
```

### C4: *OPTIONAL* Combine time series plots
We can combine the three time series plots we made so far into a single multi-panel figure. But remember, this requires that you already have these three plots available. You will also need to ensure two packages: `gtable` and `patchwork` were loaded at the beginning of this script.

```{r}
combined_timeseries <- sst_timeseries / sss_timeseries / ssf_timeseries +
  plot_layout(ncol = 1) +
  theme_minimal()

# We can check the result
combined_timeseries
```

Finally, we can save the composite figure.  

```{r}
ggsave(file.path(figure_folder, "combined_timeseries.png"), 
       combined_timeseries, width = 10, height = 12, dpi = 300)
```


## PART D: PRESSURE SECTION PLOTS
A section plot is a helpful way to visualise the water column properties along a transect, as a function of depth (y-axis) and time/location (x-axis). These plots basically show a vertical cut through the ocean as the seal experienced it during its trip. You will now create section plots for temperature, salinity, and fluorescence, which will show us the water mass properties the seal tag recorded.

First, we will prepare the raw data to create the section plots more easily. We will simply select the columns that we will need to create the plots.  

```{r}
section_data <- df |> 
  select(longitude, latitude, time, depth, pres, temp, psal, fluo_drk_npq)

head(section_data)
```

### D1: Temperature section plot
In this section, we will create a contour plot showing temperature vs depth and time. 
**Hint:** Depth should be plotted on the vertical (y) axis and the date on the horizontal (x) axis. The z axis will be whatever ocean variable you want to include in the plot.

For this example, we will use mean ocean temperature (`temp`).  

```{r}
temp_section <- section_data |>
  drop_na(temp) |> 
  ggplot(aes(x = time, y = depth, z = temp)) +
  # Add filled contours for temperature
  stat_contour_filled(bins = 10) +
  # Create date tick labels
  scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 month") +
  # TODO: Add appropriate color scale and labels
  scale_fill_cmocean("Temperature (deg C)", name = "thermal", discrete = TRUE) +
  labs(title = "Ocean temperature during seal tagging period",
       y = "Depth (m)") +
  guides(fill = guide_legend(ncol = 2)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right", 
        legend.title = element_text(hjust = 0.5, face = "italic"), 
        axis.title.x = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        plot.title = element_text(hjust = 0.5))

# Let's check the result
temp_section
```

We can now save the figure to our computer.

```{r}
ggsave(file.path(figure_folder, "temp_section.png"), temp_section, 
       width = 14, height = 8, dpi = 300)
```


### D2: Salinity section plot
Using section D1 as a reference, create a section plot for salinity.  

```{r}
# Write your code here
salt_section <- section_data |>
  drop_na(psal) |> 
  ggplot(aes(x = time, y = depth, z = psal)) +
  # Add filled contours for temperature
  stat_contour_filled(bins = 10) +
  # Create date tick labels
  scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 month") +
  # TODO: Add appropriate color scale and labels
  scale_fill_cmocean("Salinity (psu)", name = "haline", discrete = TRUE) +
  labs(title = "Ocean salinity during seal tagging period",
       y = "Depth (m)") +
  guides(fill = guide_legend(ncol = 2)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right", 
        legend.title = element_text(hjust = 0.5, face = "italic"), 
        axis.title.x = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        plot.title = element_text(hjust = 0.5))

# Let's check the result
salt_section
```


### D3: Fluorescence section plot
Create a section plot for fluorescence, but note that fluorescence does not change linearly. Instead it changes exponentially, so for the fluorescence section plot it is better to plot the log of the fluorescence values. You can do this by using the `log()` function in base `R`.

```{r}
# Write your code here
fluo_section <- section_data |>
  drop_na(fluo_drk_npq) |> 
  ggplot(aes(x = time, y = depth, z = log(fluo_drk_npq))) +
  # Add filled contours for temperature
  stat_contour_filled(bins = 10) +
  # Create date tick labels
  scale_x_date(date_labels = "%Y-%m-%d", date_breaks = "1 month") +
  # TODO: Add appropriate color scale and labels
  scale_fill_cmocean(bquote(Log~fluorescence~(mg%.%m^-3)), name = "algae", 
                     discrete = TRUE) +
  labs(title = "Ocean fluorescence during seal tagging period",
       y = "Depth (m)") +
  theme_bw() +
  guides(fill = guide_legend(ncol = 2)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        legend.title = element_text(hjust = 0.5, face = "italic"), 
        axis.title.x = element_blank(),
        panel.background = element_rect(fill = "white"),
        plot.background = element_rect(fill = "white"),
        plot.title = element_text(hjust = 0.5))

# Let's check the result
fluo_section
```


### D4: *OPTIONAL* Create another multipanel figure.

```{r}
# Write your code here
combined_section <- (temp_section+theme(axis.text.x = element_blank())) / 
  (salt_section+theme(axis.text.x = element_blank())) / fluo_section +
  plot_layout(ncol = 1)

# We can check the result
combined_section
```

